---
title: "Statistical assignment 3"
author: "Andrew Bosbery Candidate Number: 081949"
date: "13/02/2020"
output: github_document
---

```{r setup, include=FALSE}
# Please note these options.
# This tells R Markdown that we want to show code in the output document.
knitr::opts_chunk$set(echo = TRUE)
# Switching off messages in the output document. 
knitr::opts_chunk$set(message = FALSE)
# Switching on caching to make things faster (don't commit cache files on Github).
knitr::opts_chunk$set(cache = TRUE)
```

In this assignment we will explore political interest (*vote6*) and how it changes over time.

## Read data

First we want to read and join the data for the first 7 waves of the Understanding Society. (Wave 8 does not have a variable for political interest). We only want five variables: personal identifier, sample origin, sex, age and political interest. It is tedious to join all the seven waves manually, and it makes sense to use a loop in this case. Since you don't yet know about iteration I'll provide the code for you; please see the explanation of the code here: http://abessudnov.net/dataanalysis3/iteration.html.

The only thing you need to do for this code to work on your computer is to provide a path to the directory where the data are stored on your computer.

```{r}
library(tidyverse)
library(data.table)

# data.table is faster compared to readr so we'll use it in this case (the function fread()). You need to install this package first to be able to run this code.

# create a vector with the file names and paths

files <- dir(
             # Select the folder where the files are stored.
             "C:/Users/Andrew/Documents/University/Year 2/Term 2/Data Analysis 3/Project/data/UKDA-6614-tab/tab",
             # Tell R which pattern you want present in the files it will display.
             pattern = "indresp",
             # We want this process to repeat through the entire folder.
             recursive = TRUE,
             # And finally want R to show us the entire file path, rather than just
             # the names of the individual files.
             full.names = TRUE)

# Select only files from the UKHLS.
files <- files[stringr::str_detect(files, "ukhls")]
files

# create a vector of variable names
vars <- c("memorig", "sex_dv", "age_dv", "vote6")

for (i in 1:7) {
        # Create a vector of the variables with the correct prefix.
        varsToSelect <- paste(letters[i], vars, sep = "_")
        # Add pidp to this vector (no prefix for pidp)
        varsToSelect <- c("pidp", varsToSelect)
        # Now read the data. 
        data <- fread(files[i], select = varsToSelect)
        if (i == 1) {
                all7 <- data  
        }
        else {
                all7 <- full_join(all7, data, by = "pidp")
        }
        # Now we can remove data to free up the memory.
        rm(data)
} 

```

## Reshape data (20 points)

Now we have got the data from all 7 waves in the same data frame **all7** in the wide format. Note that the panel is unbalanced, i.e. we included all people who participated in at least one wave of the survey. Reshape the data to the long format. The resulting data frame should have six columns for six variables.

```{r}
Long <- all7 %>%
   gather(a_memorig:g_vote6 , key = "variable", value = "value") %>%
   separate(variable, into = c("wave", "variable"), sep = "_", extra = "merge") %>%
   spread(key = variable, value = value)
Long
```

## Filter and recode (20 points)

Now we want to filter the data keeping only respondents from the original UKHLS sample for Great Britain (memorig == 1). We also want to clean the variables for sex (recoding it to "male" or "female") and political interest (keeping the values from 1 to 4 and coding all negative values as missing). Tabulate *sex* and *vote6* to make sure your recodings were correct. 

```{r}
Long <- Long %>%
        filter(memorig == 1) %>%
        mutate(sex_dv = ifelse(sex_dv == 1, "Male", 
                               ifelse(sex_dv == 2, "Female", NA_character_)
                               )) %>%
        mutate(vote6 = case_when(
          vote6 < 0 ~ NA_integer_,
          vote6 > 0 ~ vote6
        ))

Long

Long_Sex <- Long %>%
        count(sex_dv)

Long_Sex
        
        
Long_vote <- Long %>%
        count(vote6)

Long_vote

```

## Calculate mean political interest by sex and wave (10 points)

Political interest is an ordinal variable, but we will treat it as interval and calculate mean political interest for men and women in each wave.

```{r}
meanVote6 <- Long %>%
        filter(!is.na(sex_dv)) %>%
        filter(!is.na(vote6)) %>%
        group_by(sex_dv, wave) %>%
        summarise(
                meanvote = mean(vote6, na.rm = TRUE)
        )
        
meanVote6
```

## Reshape the data frame with summary statistics (20 points)

Your resulting data frame with the means is in the long format. Reshape it to the wide format. It should look like this:

| sex_dv  | a  | b  | c  | d  |  e | f  | g  |
|---      |--- |--- |--- |--- |--- |--- |--- |
| female  |    |    |    |    |    |    |    |
| male    |    |    |    |    |    |    |    |

In the cells of this table you should have mean political interest by sex and wave.

Write a short interpretation of your findings.

```{r}
VoteWave <- meanVote6 %>%
        separate(wave, into = c("a", "b", "c", "d", "e", "f", "g")) %>%
        spread(key = "a", value = "meanvote")

VoteWave
        
```

## Estimate stability of political interest (30 points)

Political scientists have been arguing how stable the level of political interest is over the life course. Imagine someone who is not interested in politics at all so that their value of *vote6* is always 4. Their level of political interest is very stable over time, as stable as the level of political interest of someone who is always very interested in politics (*vote6* = 1). On the other hand, imagine someone who changes their value of *votes6* from 1 to 4 and back every other wave. Their level of political interest is very unstable.

Let us introduce a measure of stability of political interest that is going to be equal to the sum of the absolute values of changes in political interest from wave to wave. Let us call this measure Delta. It is difficult for me to typeset a mathematical formula in Markdown, but I'll explain this informally.

Imagine a person with the level of political interest that is constant over time: {1, 1, 1, 1, 1, 1, 1}. For this person, Delta is zero.

Now imagine a person who changes once from "very interested in politics" to "fairly interested in politics": {1, 1, 1, 1, 2, 2, 2}. For them, Delta = (1 - 1) + (1 - 1) + (1 - 1) + (2 - 1) + (2 - 2) + (2 - 2) = 1.

Now imagine someone who changes from "very interested in politics" to "not at all interested" every other wave: {1, 4, 1, 4, 1, 4, 1}. Delta = (4 - 1) + abs(1 - 4) + (4 - 1) + abs(1 - 4) + (4 - 1) + abs(1 - 4) = 3 * 6 = 18.

Large Delta indicates unstable political interest. Delta = 0 indicates a constant level of political interest.

Write the R code that does the following.

1. To simplify interpretation, keep only the respondents with non-missing values for political interest in all seven waves.
2. Calculate Delta for each person in the data set.
3. Calculate mean Delta for men and women.
4. Calculate mean Delta by age (at wave 1) and plot the local polynomial curve showing the association between age at wave 1 and mean Delta. You can use either **ggplot2** or the *scatter.smooth()* function from base R.
5. Write a short interpretation of your findings.

```{r}
LongDelta <- Long %>%
        group_by(pidp) %>%
        mutate(NumWave = length(wave)) %>%
        filter(NumWave == 7) %>%
        filter(!is.na(vote6)) %>%
        filter(!is.na(sex_dv))
        

LongDelta

Delta <- LongDelta %>%
        group_by(pidp) %>%
        summarise(Delta = sum(abs(vote6[wave == "a"] - vote6[wave == "b"]), abs(vote6[wave == "b"] - vote6[wave == "c"]), abs(vote6[wave == "c"] - vote6[wave == "d"]), abs(vote6[wave == "d"] - vote6[wave == "e"]), abs(vote6[wave == "e"] - vote6[wave == "f"]), abs(vote6[wave == "f"] - vote6[wave == "g"])))
        
        
Delta  

FullDelta <- LongDelta %>%
        left_join(Delta, by = "pidp") 
        

FullDelta

AvgDelta <- FullDelta %>%
        group_by(sex_dv) %>%
        summarise(AvgDelta =
                          mean(Delta, na.rm = TRUE))

DeltaW1 <- FullDelta %>%
        group_by(age_dv) %>%
        summarise(AvgDltaAge =
                          mean(Delta[wave == "a"]))


        
AvgDelta  

DeltaW1
        
scatter.smooth(DeltaW1,
               xlab = "Average Delta",
               ylab = "Age (Years)",
               main = "Change in Political Interest with Age")


```


From the graph we can see that from the age 50 and above there is a slight increase in the amount of people changing their political interest. Though this graph only shows there was a change and not whether it was an increase in interest or a decrease some information can be inferred. For example it suggests that some level of interaction with politics over the course of the first 30-35 years of a persons life changes their interest in politics. If there is, on average, a decrease in political interest than this suggests that as people get older they become disillusioned with politics; this means that they lose trust in politics as promises are not kept. If there is an increase in political interest it could be that as peopel get older they have more free time as they retire or earn more an so need to work less. This would allow them to take time to engage in politics and take part in it by, for example, joining a political party. 

There was no sihnificant difference between the average change in political interest for males and females. This suggests that there are few factors, or factors that are there have little effect between men and women when it comes to politics. This could mean that ideas stated above are not gender specific and that men and women are both effected in similar ways over their life. On the other hand we do not no whether the changes in political interest are positive or negative. This could mean that women become more interested in politics with age and men less interested or vice versa. Since the graph does not provide this info it would be something to remember when drawing conclusions from this graph 